### 2024-03-29 parts/starfree.tex
rewrote intro

@@

In the spirit of previous characterizations of subclasses of $\ZPoly$ in terms
of \emph{polynomial pumping arguments}
\cite{doueneau2021pebble,doueneau2022hiding,LOPEZ23b}, we will provide a
semantic property in the form of \cref{k-combinatorial:def}, that is conjecture
to characterize $\NPoly$ inside $\ZPoly$, even in the non-commutative setting.

@@

Let us now prove that the above characterizations of \kl{commutative}
\kl{$\Nat$-polyregular functions} can be combined with the recent advances in
the study of \kl{$\Rel$-polyregular functions} \cite{LOPEZ23b} to decide
aperiodicity. The key ingredient of this study is the use of a semantic
characterization of \kl{star-free $\Rel$-polyregular functions} among
\kl{$\Rel$-rational series} that generalizes the aperiodicity of languages to
functions by the means of polynomial behaviors (see
\cref{aperidic-ultimately-polynomial:ex}).

>>>>

\AP
% Connection with Weighted Automata
% introduce weighted automata and rational series here
% but whether it can be leveraged remains unclear.
Recently, aperiodicity has been shown decidable over a restricted class of
\kl{polyregular functions} where the output belongs to $\Rel$ instead of
$\Nat$, that are called \kl{$\Rel$-polyregular functions} ($\intro*\ZPoly$)
\cite{CDTL23}. These functions generalize \kl{$\Nat$-polyregular functions}
by allowing negative outputs, and therefore enabling a limited form of
backtracking. As for the unary outputs, these results generalize to
\kl{polyregular functions} having outputs in the free commutative \emph{group}
generated by a finite alphabet $\Gamma$. The decidability of aperiodicity for
\kl{$\Rel$-polyregular functions} relies on the construction of a canonical
object called the \emph{residual transducer}, the latter being essentially
based on differences between functions, crucially leveraging negative outputs
(i.e., the presence of inverse elements). Although the proof method does not carry
from $\Rel$-output functions to $\Nat$-output functions, it was conjectured
that the semantic characterization of \kl{star-free $\Rel$-polyregular
functions} ($\intro*\ZSF$) would continue to characterize \kl{star-free
$\Nat$-polyregular functions} ($\intro*\NSF$). 
% UNCLEAR

\begin{conjecture}[{\cite[Conjecture 7.61]{DOUE23}}]
    \label{zsf-nsf:conjecture}
    Let $k \in \Nat$.
    A function $f \in \NPoly[k]$
    belongs to $\NSF[k]$ if and only if
    it is \kl{ultimately polynomial} (defined page \kpageref{ultimately polynomial}).
    In particular,
    $\NSF[k] = \ZSF[k] \cap \NPoly$.
\end{conjecture}

\AP
It is well known that $\NPoly$ (resp. $\ZPoly$) is proper subclass respectively
of \intro{$\Nat$-rational series} ($\intro*\NRat$) (resp. of \intro{$\Rel$-rational
series} $\intro*\ZRat$), that have been characterized in terms of \kl{growth rate},
ambiguity, and a notion of \emph{quantitative} $\MSO$
\cite{SCHU62,KRRC13}, and therefore can benefit from generic
theorems obtained on these \cite{BERE10}. However, for these
larger classes of functions, several competing notions of aperiodicity exist
\cite{DRGA19,REUT80,CDTL23}, and the
decidability of the ones that extends $\ZSF$ and $\NSF$ remains open.
Furthermore, the decidability of $\NRat$ inside $\ZRat$ is also a longstanding
open problem, that was to the knowledge of the authors, only solved by
\cite{KARH77} in the case of \kl{$\Rel$-rational series} that compute
\emph{polynomials}, with an incorrect proof. Furthermore, decidability of
$\NPoly$ inside $\ZPoly$ (resp. $\NSF$ inside $\ZSF$) remains an open problem
\cite[Open question 5.55]{DOUE23}.

@@

In the spirit of previous characterizations of subclasses of $\ZPoly$ in terms
of \emph{polynomial pumping arguments}
\cite{DOUE21,DOUE22,CDTL23}, we will provide a
semantic property in the form of \cref{k-combinatorial:def}, that is conjectured
to characterize $\NPoly$ inside $\ZPoly$, even in the non-commutative setting.

@@

Let us now prove that the above characterizations of \kl{commutative}
\kl{$\Nat$-polyregular functions} can be combined with the recent advances in
the study of \kl{$\Rel$-polyregular functions} \cite{CDTL23} to decide
aperiodicity. The key ingredient of this study is the use of a semantic
characterization of \kl{star-free $\Rel$-polyregular functions} among
\kl{$\Rel$-rational series} that generalizes the aperiodicity of languages to
functions by the means of polynomial behaviors (see
\cref{aperidic-ultimately-polynomial:ex}).

### 2024-03-29 parts/transducers.tex
update ICALP
rewrote intro

\AP The goal of this section is to provide tools that do not require the
assumption of \kl{commutativity}. To that end, we introduce the notion of
\kl{$k$-residual transducer}, that generalizes the notion of residual
transducer introduced for \kl{$\Rel$-polyregular functions} in \cite{LOPEZ23b}.
This notion of residual transducer has its own interest, as it shifts the
attention from (compatible) equivalence relations of finite index that are the
classical tool in automata theory, towards (compatible) order relations that
are \kl{well-quasi-ordered}, which is the order counterpart of having finite
index. Let us recall that a sequence $\seqof{u_i}{i \in \Nat}$ of elements in a
quasi-ordered set $(X, \leq)$ is \intro{good} whenever there exist $i < j$ such
that $u_i \leq u_j$. The set $X$ is a \intro{well-quasi-ordering} when every
infinite sequence is \reintro{good}. A sequence is \intro{bad} when it is not
\reintro{good}. The notion of \reintro{good sequences} can also be applied to
binary relations that are not orderings, and a binary relation $R$ for which
every infinite sequence is \reintro{good} is said to be \intro{well}
\cite{MELL98}.

As a model of computation, we will use the following definition of transducers,
that is based on the marble transducers \cite{EHB99}, and follows the general
pattern of transducers introduced in \cite{LOPEZ23b}. Note that we only
consider $\Nat$-valued functions, but could consider other commutative
semigroups.

@@

\begin{lemma}[{\cite[Theorems 5.15 and 7.10]{gaetanphd}}]
    \label{transducer-nsf-npoly:lemma}
    Let $f \colon \Sigma^* \to \Nat$, and let $k \in \Nat$.
    The following are equivalent:
    \begin{itemize}
        \item $f \in \NPoly[k]$ (respectively $\NSF[k]$),
        \item $f$ is \kl{computed} by an \kl{$\NPoly[k-1]$-transducer}
            (respectively by an \kl{$\NSF[k-1]$-transducer} that is 
            \kl{counter-free}).
    \end{itemize}
\end{lemma}

\AP The core concept of this section is that of a \intro{residual} of a
function $f \colon \Sigma^* \to \Rel$,  defined by $\intro*\app{f}{u} \defined
w \mapsto f(uw)$. The collection of \reintro{residuals} of a function $f$ is
denoted $\intro*\Res(f)$ and is defined as the set of $\reintro*\app{f}{u}$
where $u$ ranges over words in $\Sigma^*$. Let $k \in \Nat$. We define $f
\intro*\zpolyequiv[k] g$ if and only if $(g - f) \in \ZPoly[k-1]$, and $f
\intro*\npolyleq[k] g$ if and only if $(g - f) \in \NPoly[k-1]$. Given a
function $f \colon \Sigma^* \to \Rel$, we are going to study both $(\Res(f),
\reintro*\zpolyequiv[k])$, and $(\Res(f), \reintro*\npolyleq[k])$. However, to
simplify notations, instead of writing $\app{f}{u} \reintro*\npolyleq[k]
\app{f}{v}$ when $u,v \in \Sigma^*$, we will use the convenient notation $u
\intro*\resleq{f}{k} v$, and directly consider the space $(\Sigma^*,
\reintro*\resleq{f}{k})$.

>>>>

\AP The goal of this section is to provide tools that do not require the
assumption of \kl{commutativity}. To that end, we introduce the notion of
\kl{$k$-residual transducer}, that generalizes the notion of residual
transducer introduced for \kl{$\Rel$-polyregular functions} in \cite{CDTL23}.
This notion of residual transducer has its own interest, as it shifts the
attention from (compatible) equivalence relations of finite index that are the
classical tool in automata theory, towards (compatible) order relations that
are \kl{well-quasi-ordered}, which is the order counterpart of having finite
index. Let us recall that a sequence $\seqof{u_i}{i \in \Nat}$ of elements in a
quasi-ordered set $(X, \leq)$ is \intro{good} whenever there exist $i < j$ such
that $u_i \leq u_j$. The set $X$ is a \intro{well-quasi-ordering} when every
infinite sequence is \reintro{good}. A sequence is \intro{bad} when it is not
\reintro{good}. The notion of \reintro{good sequences} can also be applied to
binary relations that are not orderings, and a binary relation $R$ for which
every infinite sequence is \reintro{good} is said to be \intro{well}
\cite{MELL98}.

As a model of computation, we will use the following definition of transducers,
that is based on the marble transducers \cite{EHB99}, and follows the general
pattern of transducers introduced in \cite{CDTL23}. Note that we only
consider $\Nat$-valued functions, but could consider other commutative
semigroups.

@@

\begin{lemma}[{\cite[Theorems 5.15 and 7.10]{DOUE23}}]
    \label{transducer-nsf-npoly:lemma}
    Let $f \colon \Sigma^* \to \Nat$, and let $k \in \Nat$.
    The following are equivalent:
    \begin{itemize}
        \item $f \in \NPoly[k]$ (respectively $\NSF[k]$),
        \item $f$ is \kl{computed} by an \kl{$\NPoly[k-1]$-transducer}
            (respectively by an \kl{$\NSF[k-1]$-transducer} that is 
            \kl{counter-free}).
    \end{itemize}
\end{lemma}

\AP The core concept of this section is that of a \intro{residual} of a
function $f \colon \Sigma^* \to \Rel$,  defined by $\intro*\app{f}{u} \defined
w \mapsto f(uw)$. The collection of \reintro{residuals} of a function $f$ is
denoted $\intro*\Res(f)$ and is defined as the set of $\reintro*\app{f}{u}$
where $u$ ranges over words in $\Sigma^*$. Given $k \in \Nat$, we define $f
\intro*\zpolyequiv[k] g$ if and only if $(g - f) \in \ZPoly[k-1]$, and $f
\intro*\npolyleq[k] g$ if and only if $(g - f) \in \NPoly[k-1]$. The spaces of
interest for a function $f \in \ZPoly[k]$ will be $(\Res(f),
\reintro*\zpolyequiv[k])$, and $(\Res(f), \reintro*\npolyleq[k])$. However, to
simplify notations, instead of writing $\app{f}{u} \reintro*\npolyleq[k]
\app{f}{v}$ when $u,v \in \Sigma^*$, we will use the convenient notation $u
\intro*\resleq{f}{k} v$, and directly consider the space $(\Sigma^*,
\reintro*\resleq{f}{k})$.
